---
title: "Assignment 5"
author: "Jieran Sun, Hui Jeong (HJ) Jung, Gudmundur Björgvin Magnusson"
output: pdf_document
date: "2023-03-28"
---

## Question 12

### (1)

```{=tex}
\begin{align}
  P(t + dt) = P(dt)P(t) = (I + Rdt)P(t)
\end{align}
```
Therefore we can show that

```{=tex}
\begin{align}
  \frac{dP(t)}{dt} = RP(t)
\end{align}
```
### (2)

As the Markov chain is homogeneous Markov chain and $\pi$ is the ergodic stationary distribution,

```{=tex}
\begin{align}
  t \rightarrow \infty \ \  P(t) \pi &= \pi     \\
  \text{multiply $R$,} \ \ R P(t) \pi &= R \pi  \\
  \frac{dP(t)}{dt} \pi &= R \pi \\
\end{align}
```
Given that when $t \rightarrow \infty$, $P(t)$ reach steady state, $\frac{dP(t)}{dt} = 0$. Hence $R \pi = 0$.

## Question 13

### (1) The joint probability of the tree is

```{=tex}
\begin{align}
  P(X,Z|T) = P(Z_4) * P(X_5|Z_4) * P(Z_3|Z_4) * P(Z_2|Z_3) * P(Z_1|Z_3) \\
  * P(X_4|Z_2) * P(X_3|Z_2) * P(X_2|Z_1) * P(X_1|Z_1)
\end{align}
```
### (2) To do the naive calculation of P(X\|T) via brute-force marginalization over the hidden nodes Z,

for each node X, we have to marginalize out all the internal nodes Z that it is dependent on. Here is an example with X4.

```{=tex}
\begin{align}
  P(X_4|T) &= \sum_Z P(X_4, Z|T) = \sum_{Z_2, Z_3, Z_4} P(X_4|Z_2)P(Z_2|Z_3)P(Z_3|Z_4)P(Z_4)\\
\end{align}
```
Given each Z represent a nucleitode and can take 4 values(A,T,G,C), then the total number of cases for $X_4$ is $4^3 = 64$. Hence 64 summation operations are needed for X4.

Similarly, for $X_1, X_2, X_3$, they all have 64 cases and 64 summation operations, and for $X_5$ we have 4 summation cases as $P(X_5, Z) = \sum_{Z_4} P(X_5 | Z_4)P(Z_4)$ and it only has 4 cases.

Hence in total the number of summation operation is $64 \times 4 + 4 = 260$.

### (3)

```{=tex}
\begin{align}
  P(X,Z|T) = P(Z_4) * P(X_5|Z_4) * P(Z_3|Z_4) * P(Z_2|Z_3) * P(Z_1|Z_3) \\
  * P(X_4|Z_2) * P(X_3|Z_2) * P(X_2|Z_1) * P(X_1|Z_1)
\end{align}
```
By rearranging the expression in such way we only need to do 4 summations.

## Question 14

### (1)

```{r setup}
if(!require("phangorn")) {
  install.packages("phangorn")
}
if(!require("ape")) {
  install.packages("ape")
}

library(phangorn)
library(ape)
```

```{r load data}
ParisRT <- read.dna("ParisRT.txt")
```

### (2)

```{r initial tree}
distParis <- dist.dna(ParisRT)
initTree <- NJ(distParis)
plot(initTree)
```

### (3)

```{r fit Kimura}
kimura <- pml(tree= initTree, data= phyDat(ParisRT), model= "K80")
kimura$logLik
```

The log likelihood of the fitted model is -3003.487.

### (4)

The values of the optimised rate matrix can be found below.

```{r optimise parameters}
optimParam <- optim.pml(kimura, optQ= TRUE)
optimParam
```

### (5)

After optimizing with respect to branch lengths, nucleotide substitution rates, and tree topology, the results are as below.

```{r optimise more parameters}
optimParam2 <- optim.pml(kimura, optQ = TRUE, optNni = TRUE, optEdge = TRUE)
optimParam2$logLik
```

The log likelihood is -2849.789.

### (6)

```{r bootstrap, results ='hide'}
bootPML <- bootstrap.pml(optimParam2, optNni= TRUE)
```

In this bootstrapping function, we are resampling the sequences that we are using to construct a phylogenetic tree, and seeing if the same branch is observed even when generating a new tree based on bootstrapped data, which would indicate confidence in the observed branch.

### (7)

```{r plot tree}
plotBS(tree= optimParam2$tree, BStrees = bootPML, type= "phylogram")
```

Judging from this plot, it is more likely that Mme_S was more likely to have infected the patient Mme_L.
