---
title: "Assignment 5"
author: "Jieran Sun, Hui Jeong (HJ) Jung, Gudmundur Björgvin Magnusson"
output: pdf_document
date: "2023-03-28"
---

## Question 12

### (1)
```{=tex}
\begin{align}
  P(t + dt) = P(dt)P(t) = (I + Rdt)P(t)
\end{align}
```
Therefore we can show that 
```{=tex}
\begin{align}
  \frac{dP(t)}{dt} = RP(t)
\end{align}
```

### (2)
## TODO 

## Question 13

### (1) The joint probability of the tree is 
## TODO : not sure what the "long version" is supposed to be, bc the one below is the "simple" one 
```{=tex}
\begin{align}
  P(X,Z|T) = P(Z_4) * P(X_5|Z_4) * P(Z_3|Z_4) * P(Z_2|Z_3) * P(Z_1|Z_3) \\
  * P(X_4|Z_2) * P(X_3|Z_2) * P(X_2|Z_1) * P(X_1|Z_1)
\end{align}
```

### (2) To do the naive calculation of P(X|T) via brute-force marginalization over the hidden nodes Z, 
for each node X, we have to marginalize out all the internal nodes Z that it is dependent on. 
Here is an example with X4. 
```{=tex}
\begin{align}
  P(X_4|T) = \sum_Z P(X_4, Z|T) = \sum_Z P(X_4|Z_2)P(Z_2|Z_3)P(Z_3|Z_4)P(Z_4)\\
  = \sum_Z P(X_4, Z_2, Z_3, Z_4)
\end{align}
```
Thus for X4, we would have to sum over 3 times. For X1, X2, X3, and X4 they all depend on 3 Zs so they also have to sum 3 times. 
X5 only depends on Z4, therefore only has to sum once. In total the brute force method would require 13 summations.  

### (3)
```{=tex}
\begin{align}
  P(X,Z|T) = P(Z_4) * P(X_5|Z_4) * P(Z_3|Z_4) * P(Z_2|Z_3) * P(Z_1|Z_3) \\
  * P(X_4|Z_2) * P(X_3|Z_2) * P(X_2|Z_1) * P(X_1|Z_1)
\end{align}
```
By rearranging the expression in such way we only need to do 4 summations. 

## Question 14

### (1)
```{r setup}
if(!require("phangorn")) {
  install.packages("phangorn")
}
if(!require("ape")) {
  install.packages("ape")
}

library(phangorn)
library(ape)
```

```{r load data}
ParisRT <- read.dna("ParisRT.txt")
```

### (2)
```{r initial tree}
distParis <- dist.dna(ParisRT)
initTree <- NJ(distParis)
plot(initTree)
```
### (3)
```{r fit Kimura}
kimura <- pml(tree= initTree, data= phyDat(ParisRT), model= "K80")
kimura$logLik
```
The log likelihood of the fitted model is -3003.487. 

### (4)
The values of the optimised rate matrix can be found below. 
```{r optimise parameters}
optimParam <- optim.pml(kimura, optQ= TRUE)
optimParam
```
### (5)
After optimizing with respect to branch lengths, nucleotide substitution rates, and tree topology, 
the results are as below. 
```{r optimise more parameters}
optimParam2 <- optim.pml(kimura, optQ = TRUE, optNni = TRUE, optEdge = TRUE)
optimParam2$logLik
```
The log likelihood is -2849.789.

### (6)
```{r bootstrap, results ='hide'}
bootPML <- bootstrap.pml(optimParam2, optNni= TRUE)
```
In this bootstrapping function, we are resampling the sequences that we are using to construct
a phylogenetic tree, and seeing if the same branch is observed even when generating a new tree
based on bootstrapped data, which would indicate confidence in the observed branch. 

### (7)
```{r plot tree}
plotBS(tree= optimParam2$tree, BStrees = bootPML, type= "phylogram")
```
Judging from this plot, it is more likely that Mme_S was more likely to have infected the patient Mme_L. 