---
title: "Assignment 9"
output: pdf_document
date: "2023-04-29"
author: Hui Jeong (HJ) Jung, Gudmundur Björgvin Magnusson, Jieran Sun
---

```{r setup, include=FALSE}
library(ggplot2)
library(pcalg)
library(BiDAG)
library(igraph)
library(pheatmap)
```

## Problem 24 Testing for marignal correlation 

First we load the data and examine the data structure.
```{r load data }
nvmData <- readRDS("MVN_DAG.rds")
head(nvmData)
```

Plot the scatter plot for the observation A and B
```{r scatter plots}
ggplot(nvmData, aes(x = A, y = B)) + geom_point() +
  ggtitle("Scatter plot between observation A and B")
```

It can be inferred from the scatter plot that the correlation between observation A and B are merely nonexistent, A and B are not correlated with each other. This observation from the scatter plot also agree with what we have observed on the DAG graph where A and B are independent of each other. 

Now we test the independence hypothesis using `cor.test()`

```{r cortest marginal correlation}
cor.test(nvmData$A, nvmData$B)
```

Since the p value is way larger than 0.05, we accept the null hypothesis and confirm that A and B are independent from each other.

## 25 partial correlation

First, we find the residual of A and B by fitting them into a linear regression on C
```{r finding residual}
Ares <- residuals(lm(A ~ C, data = nvmData)) 
Bres <- residuals(lm(B ~ C, data = nvmData)) 
```

Then similarly we plot the residual relationship

```{r scatter plot residuals}
ggplot() + geom_point(aes(x = Ares, y = Bres)) +
  ggtitle("residual relationship between A and B")
```
Comparing to the results shown in question 24, this plots show us a stronger correlation between the residual of A and B observations, indicating that A and B are not conditionally independent given C.

Now let's test the null hypothesis that A and B are conditionally independent given C.
```{r partial correlation test}
cor.test(Ares, Bres)
```

Given that p value is way smaller than 0.05, we reject the null hypothesis and state that A and B are not conditionally independent given C. This result also agree with what we observed from Figure 2.

## 26 run PC algorithm 

Given the instruction, we run the pc function and plot the resulting DAG
```{r run the pc algorithm}
statList <- list(C = cor(nvmData), n = nrow(nvmData))

pcResults <- pc(suffStat = statList, 
                indepTest = gaussCItest, 
                alpha = 0.9,
                labels = colnames(nvmData))

iplotPC(pcResults)
```
## 27 partition MCMC algorithm

First we set up the score and run the MCMC algorithm
```{r set up the s}
# initialize the parameters 
score <- scoreparameters("bge", nvmData)

# run iterative MCMC algorithm
maxBNalpha05 <- learnBN(score, algorithm = "orderIter", alpha = 0.05)
# plot the maxBN DAG results
plot.igraph(graph_from_adjacency_matrix(maxBNalpha05$CPDAG, mode= "directed"),
            main = "alpha = 0.05")

# set alpha to 0.1
maxBNalpha10 <- learnBN(score, algorithm = "orderIter", alpha = 0.1)
# plot the maxBN DAG results
plot.igraph(graph_from_adjacency_matrix(maxBNalpha10$CPDAG, mode= "directed"), 
            main = "alpha = 0.1")

# set alpha to 0.2
maxBNalpha50 <- learnBN(score, algorithm = "orderIter", alpha = 0.5)
# plot the maxBN DAG results
plot.igraph(graph_from_adjacency_matrix(maxBNalpha50$CPDAG, mode= "directed"), 
            main = "alpha = 0.5")
```
From the list of graphs above we can see the hyperparameter alpha does not affect the final DAG connectivity and shape.

Then we run the MCMC partition algorithm, extract the marginal probability for the edges and plot them into a heatmap
```{r partition MCMC algorithm}
partitionSample <- sampleBN(score, 
                            algorithm = "partition",
                            startspace = maxBNalpha05$endspace)

# compute the marginal posterior probabilities
edgesPosterior <- edgep(partitionSample, pdag = TRUE)

# plot the heatmap 
pheatmap(edgesPosterior)
```


