---
title: "Assignment 7"
output: pdf_document
date: "2023-04-08"
author: Hui Jeong (HJ) Jung, Gudmundur Björgvin Magnusson, Jieran Sun
---

```{r setup, include=FALSE}
library(mnem)
library(dplyr)
library(data.table)
library(ggplot2)
```

## Problem 21 Hidden Markov NEMs

1. Implement Hidden markov model in the transitional probability between the graphs. 

```{r set up the function for HM-NEMs algorithm}
adjMatU <- matrix(c(1,1,1,0,
                    0,1,1,1,
                    0,0,1,1,
                    0,0,0,1),
                  nrow = 4, byrow = TRUE)

adjV1 <- matrix(c(1,1,1,0,
                  0,1,1,1,
                  0,0,1,0,
                  0,0,0,1),
                nrow = 4, byrow = TRUE)

adjV2 <- matrix(c(1,0,0,0,
                  1,1,1,0,
                  1,0,1,0,
                  1,0,0,1),
                nrow = 4, byrow = TRUE)

endAdj <- list(adjV1, adjV2)

transitionProb <- function(startAdj, endAdjs, lambda){
  transProb <- sapply(endAdjs, function(endAdj){
    Suv <- sum(abs(endAdj - startAdj))
    Tuv <- lambda * ((1 - lambda)^Suv)
  })
  transPorb <- transProb/sum(transProb)
  
  return(transPorb)
}

transProb <- lapply(seq(0.1, 0.9, by = 0.1), function(lambda){
  transitionProb(startAdj = adjMatU, endAdjs = endAdj, lambda = lambda)
}) %>% as.data.frame() %>% t() %>% as.data.table()

transProb[, lambda := seq(0.1, 0.9, by = 0.1)]

```

```{r print out the results}
knitr::kable(transProb)
```

2. Plot out the results

```{r plot out the results}
ggplot(data = transProb) + 
  geom_line(aes(x = lambda, y = V1, col = "V1")) + 
  geom_line(aes(x = lambda, y = V2, col = "V2"))
```

## Problem 22 

1. Given the circumstances, the perturbation map rho

```{r print out the perturbation map}

rho <- matrix(c(1,0,1,0,
                0,1,1,1),
              nrow = 2, byrow = TRUE)

print("rho: ")
print(rho)
```

2. (a) Given the graphs shown in F1 and F2, the expected effect pattern for both cases are

```{r F for both cases}
FList <- list()

FList[["K1"]] <- matrix(c(1,0,1,0,
                          1,1,1,1), nrow = 2, byrow = TRUE)
FList[["K2"]] <- matrix(c(0,1,1,1,
                          1,1,1,1), nrow = 2, byrow = TRUE)

FList
```

2 (b) Given c1 and c2 are from F1, and c3 and c4 are from F2. The final R matrix is 

```{r}
RMat <- as.matrix(cbind(FList$K1[,c(1,2)],
                        FList$K2[,c(3,4)]))
RMat[RMat == 0 ] <- -1 
RMat
```


3 Calculating the EM algorithm to update the weight matrix

```{r calculate the loglikelihood in each case}
logList <- list()
logList[["K1"]] <- diag(t(FList$K1) %*% RMat)
logList[["K2"]] <- diag(t(FList$K2) %*% RMat)
logList
```

Then we calculate the responsibility
```{r calculate the log-likelihood }
pi <- c(0.44, 0.56)

resp <- rbind(pi[1]*exp(logList$K1),
              pi[2]*exp(logList$K2)) %>% as.data.table()

resp <- resp[, lapply(.SD, function(x) x/sum(x)), .SD = colnames(resp)]
resp
```
Based on the responsibility, we update the weight as 
```{r update weight}
(newWeight <- rowSums(resp)/sum(resp))
```
